Towards Verifying the Bitcoin-S Library

### 2019-07-05 paper/paper.tex
typos etc

\begin{figure}
\begin{lstlisting}[style=scala]
  def factorial(n: Int): Int = {
      require(n >= 0)
      if (n == 0) {
        1
      } else {
        n * factorial(n - 1)
      }
  } ensuring(res => res >= 0)
\end{lstlisting}
	\caption{Factorial function with specification}
	\label{fig:factorial}
\end{figure}
Our function does not satisfy the specification. An overflow in the
32-bit Int leads to a negative result for the input 17, as Stainless
reports in Figure~\ref{fig:failed}. Changing the type from Int to
BigInt will result in a successful verification.

@@

What will turn out to be more relevant for us are the Scala features
which Stainless does not support, such as: (concrete) class
definitions, inheritance by objects, abstract type members, and inner
classes in case objects.

@@

Now that we have a transaction output, we create a transaction to
spend it in Figure~\ref{fig:tx}. First, we need a reference to an
output of a previous transaction, here called
\texttt{outPoint}. Second, we add some information on how to spend
that output, in particular, how to sign the transaction to allow
this. Now we assemble the list of unspent transaction outputs (utxos),
in our case just one.

@@

We then create a lock script (\texttt{destinationSPK}) to receive the
coins, create our list of transaction outputs (\texttt{destinations}),
define the fee rate and set some bitcoin network parameters.

@@

\paragraph{A Bug in the checkTransaction Function.} Note lines 15-17
of the checkTransaction function in Figure~\ref{fig:checktrans}. Here,
value prevOutputTxIds gathers a list of all transaction indentifiers
referenced by the inputs of the current transaction. If the size of
this list is the same as the size of this list with duplicates
removed, we know that no transaction has been referenced twice. This
prevents a transaction from spending two different outputs of the same
previous transaction. The check is too strict: checkTransaction
returns false for valid transactions.

The fix is simple: we perform the duplicate check on the
TransactionOutPoints instead of on their transaction identifiers.  A
TransactionOutPoint contains the txId as well as the output index it
references. Note that TransactionOutPoint is a case class and thus has
a built in == method.

@@

It is of course a crucial property we are verifying here: if zero
satoshis were credited to your account, you would not want your
balance to change! It is also the simplest meaningful property to
verify that we can think of. However, the code involved in performing
the addition of two satoshi amounts in bitcoin-s is non-trivial. The
reason for that is a pecularity of consensus code: agreement with the
majority is more important than correctness, whatever correctness
might mean. The most widely used bitcoin implementation by far is the
reference implementation Bitcoin Core, written in C++. For consensus
code, bitcoin-s has little choice but to be in strict agreement with
the reference implementation. To achieve that, it implements C-like
data types in Scala and then implements functionality using those
C-like data types. For example, the Satoshis class, which represents
an amount of satoshis, is implemented using the class Int64 which
aims to represent the C-type \texttt{int64\_t}.

@@

\paragraph{Inheriting Objects.} In both files we have objects
extending the BaseNumbers trait, on lines 30 and 23 respectively,
which Stainless does not support. We simply turn those onjects into
case objects. That transformation is equivalent. Case objects have
various additional properties (for example, being serializable) but
none of our code depends on the absence of any of that.

@@

\paragraph{Self-Reference in Type Parameter Bound.} In
NumberTypes.scala on lines 3 and 19 are classes a type parameter and a
type boundary that contains that typarameter itself. Stainless does
not currently support such self-referential type boundaries. We opened
an issue \cite{Stainless:issue519} on the Stainless repository and the
developers have targeted version 0.4 to support self-referential type
boundaries. Since our code only uses Number with type parameter T
instantiated to Int64, we just remove the type parameter declaration
and replace all its occurrences it by Int64.

\paragraph{Missing Member \texttt{bigInteger} in BigInt.} In NumberType
on line 6 there is a reference to \texttt{bigInteger}. The Scala
BigInt class is essentially a wrapper around
java.math.BigInteger. BigInt has a member bigInteger which is the
underlying instance of the Java class. The Java class has a method
longValueExact which returns a long only if the BigInteger fits into a
long, otherwise throws exception. Stainless does not support Java
classes and in particular its BigInt has no member
bigInteger. However, our code never calls toLong anymore, so we just
remove it.

\paragraph{Type Members.} In NumberType.scala there is a type member
on line 4. Our version of Stainless (0.1) does not support type
members. We just remove the declaration and replace all occurrences of
A with BigInt, since A is never overwritten in an implementing class.
Note that in the mean time Stainless has implemented support for type
members \cite{Stainless:pull470}.  Since version 0.2 verification
should succeed without this change.

@@

On the other hand, bitcoin-s uses features that might not be supported
even by future Stainless versions, such as (concrete, non-case)
classes \todo{is that even true? where?} or calls to Java code. Here
the bitcoin-s code can hopefully be adapted to accomodate formal
verification.

From our results we conclude that formal verification of bitcoin
libraries in general and bitcoin-s in particular is a worthwile
endeavour. We are looking forward to verifying more substantial parts
of the code in future work.

@@

\todo{reference to verified code}

>>>>

\begin{figure}
\begin{lstlisting}[style=scala]
  def factorial(n: Int): Int = {
      require(n >= 0)
      if (n == 0) {
        1
      } else {
        n * factorial(n - 1)
      }
  } ensuring(res => res >= 0)
\end{lstlisting}
	\caption{Factorial function with specification}
	\label{fig:factorial}
\end{figure}
Our function does not satisfy the specification. An overflow in the
32-bit Int leads to a negative result for the input 17, as Stainless
reports in Figure~\ref{fig:failed}. Changing the type Int to BigInt
will result in a successful verification.

@@

What will turn out to be more relevant for us are the Scala features
which Stainless does not support, such as: arbitrary class
definitions, inheritance by objects, abstract type members, and inner
classes in case objects.

@@

Now that we have a transaction output, we create a transaction to
spend it in Figure~\ref{fig:tx}. First, we need a reference to an
output of a previous transaction, here called
\texttt{outPoint}. Second, we add some information on how to spend
that output, in particular, how to sign the transaction. Now we
assemble the list of unspent transaction outputs (utxos), in our case
just one.

@@

We then create a lock script (\textit{destinationSPK}) to receive the
coins, create our list of transaction outputs (\texttt{destinations}),
define the fee rate and set some bitcoin network parameters.

@@

\paragraph{A Bug in the checkTransaction Function.} Note lines 15-17
of the checkTransaction function in Figure~\ref{fig:checktrans}. Here,
the value prevOutputTxIds gathers a list of all transaction identifiers
referenced by the inputs of the current transaction. If the size of
this list is the same as the size of this list with duplicates
removed, we know that no transaction has been referenced twice. This
prevents a transaction from spending two different outputs of the same
previous transaction. The check is too strict: checkTransaction
returns false for valid transactions.

The fix is simple: we perform the duplicate check on the
TransactionOutPoint instances instead of on their transaction
identifiers. Note that TransactionOutPoint is a case class and thus
its notion of equality is just what we need: equality of of both the
txId and the output index.

@@

It is of course a crucial property we are verifying here: if zero
satoshis were credited to your account, you would not want your
balance to change! It is also the simplest meaningful property to
verify that we can think of. However, the code involved in performing
the addition of two satoshi amounts in bitcoin-s is non-trivial. The
reason for that is a peculiarity of consensus code: agreement with the
majority is more important than correctness, whatever correctness
might mean. The most widely used bitcoin implementation by far is the
reference implementation Bitcoin Core, written in C++. For consensus
code, bitcoin-s has little choice but to be in strict agreement with
the reference implementation. To achieve that, it implements C-like
data types in Scala and then implements functionality using those
C-like data types. For example, the Satoshis class, which represents
an amount of satoshis, is implemented using the class Int64 which
aims to represent the C-type \texttt{int64\_t}.

@@

\paragraph{Inheriting Objects.} In both files we have objects
extending the BaseNumbers trait, on lines 30 and 23 respectively,
which Stainless does not support. We simply turn those objects into
case objects. That transformation is equivalent. Case objects have
various additional properties (for example, being serializable) but
none of our code depends on the absence of those.

@@

\paragraph{Self-Reference in Type Parameter Bound.} In
NumberTypes.scala both on lines 3 and 19 is a class with a type
parameter and a type boundary that contains that type parameter
itself. Stainless does not currently support such self-referential
type boundaries. We opened an issue \cite{Stainless:issue519} on the
Stainless repository and the developers have targeted version 0.4 to
support self-referential type boundaries. Since our code only uses
Number with type parameter T instantiated to Int64, we just remove the
type parameter declaration and replace all its occurrences by Int64.

\paragraph{Missing Member \texttt{bigInteger} in BigInt.} In
NumberType on line 6 there is a reference to \texttt{bigInteger}. The
Scala BigInt class is essentially a wrapper around
java.math.BigInteger. BigInt has a member bigInteger which is the
underlying instance of the Java class. The Java class has a method
longValueExact which returns a long only if the BigInteger fits into a
long, otherwise throws exception. Stainless does not support Java
classes and in particular its BigInt has no member
bigInteger. However, our code does not call toLong anymore, so we just
remove it.

\paragraph{Type Members.} In NumberType.scala there is a type member
on line 4. Our version of Stainless (0.1) does not support type
members. We just remove the declaration and replace all occurrences of
A with BigInt, since A is never overwritten in an implementing class.
Note that in the meantime Stainless has implemented support for type
members \cite{Stainless:pull470}.  Since version 0.2 verification
should succeed without this change.

@@

On the other hand, bitcoin-s uses features that might not be supported
even by future Stainless versions, such as (concrete, non-case)
classes \todo{NOT TRUE, no classes found} or calls to Java code. Here
the bitcoin-s code can hopefully be adapted to accommodate formal
verification.

From our results we conclude that formal verification of bitcoin
libraries in general and bitcoin-s in particular is a worthwhile
endeavour. We are looking forward to verifying more substantial parts
of the code in future work.

@@

\todo{reference to verified code} \todo{say which version of stainless
  we used, which version of bitcoin-s:
  91633375a37a170ba260ff4b443582aca4aa0c77 }

\todo{code in text: italic or courier?}
